Accounting for sampling error in when applying matrix factorization to gene expression data

### 2021-01-24 Methods_Paper.tex
starting to slim it down...

\documentclass[tikz, 11pt,a4paper,oneside,fleqn]{article}

\usepackage[margin=2.5cm]{geometry}
%\usepackage
%[
%        left=3cm,
%        right=2cm,
%        top=2cm,
%        bottom=3cm,
%        % use vmargin=2cm to make vertical margins equal to 2cm.
%        % us  hmargin=3cm to make horizontal margins equal to 3cm.
%        % use margin=3cm to make all margins  equal to 3cm.
%]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{setspace}
\usepackage{enumitem}
\setlist[description]{leftmargin=\parindent,labelindent=\parindent}

@@

\title{MF Methods paper}
\author{Ian Poole and Ailith Ewing}
\date{January 2021}

@@

\subsection{Epithelial Ovarian Cancer}
\label{sec-epithelial-ovarian-cancer}

@@

\subsection{Patterns of gene expression in HGSOC}
\label{sec-review-of-gene-expression}
Identifying the patterns -- or ``signatures'' -- of gene expression has been an active research area, for the purposes of understanding the carcinogenesis pathway, subtype classification, predicting survival prognosis, predicting treatment response and understanding the causative mechanism of risk factors; there is much cross-over between these areas, however.

\subsubsection{Carcinogenesis pathways including homologous repair deficiency (HRD)}

BRCA1 inactivation is known to cause chromosomal instability in many cancers.  Pradhan \etal\ \cite{Pradhan2010} investigated the role of BRCA1 in HGSOC by copy number and expression analysis, finding surprisingly that inactivation has no relationship with gross genomic alteration.  They leave open the question whether DNA repair by PARP plays a role.   The relationship between BRCA1/2 and DNA repair -- specifically homologous repair deficiency -- is picked up by Ewing \etal\ \cite{Ewing2020}.  They study the complex interplay of single nucleotide variants and large structural variants as they impact BRCA1/2 disruption and thence HRD.  This work suggests that when BRCA1/2 loss is detected in HGSOC patients, there may be a clinical role for PARP inhibitors, since this would prevent error-prone non-homologous repair by PARP, and so selectively kill BRCA1/2 disrupted cells.

A gene expression signature of HRD has been reported \cite{Lu2014} and shown to be predictive of overall and progression free survival in OC patients.  The signature was derived by a supervised approach, from tumour samples known to be with or without HRD.  Examining the paper's supplementary material, the signature is based on 116 genes, with +1 /-1 weights.  Notable positively weighted genes include ABCF1, ACTL6B and ARID5A; negatively weighted include ZNF780A, WDR36 and TRPM7.

Studying the combined landscape of chromosomal rearrangements, driver mutations (particularly TP53, BRCA1/2), methylation and gene expression profiles has lead to a better understanding of the molecular events involved in the progression HGSOC and the development of chemoresistance \cite{Patch2015}.  The sequence of events is complex, however, and not easily summarised here.   A similar integrative / multiplatform approach is taken in \cite{He2017a} and \cite{Hoadley2014}.

\subsubsection{Subtype classification}
\label{sec-subtype-classification}

A relatively early (2003) expression microarray based work demonstrated clustering into lymphocyte related genes (IGH3, IGKC, IGLJ3), extracellular matrix/stromal related (COL11A1, COL3A1, MMP2, SPARC, RBBP1) and six other clusters \cite{Schaner2003}.

Wang \etal\ \cite{Wang2017c} use NMF clustering (see section \ref{sec-nmf}) to identify five subtypes of HGSOC informative of outcomes: 1:mesenchymal, 2:immunoreactive, 3:proliferative, 4:differentiated and 5:anti-mesenchymal.
Subtypes 2 and 5 are found to be associated with better survival.

Disruption of the PI3K-AKT signalling pathway is known to be significant in several cancers, due to its role in regulating apoptosis.  
Espinosa \etal\ \cite{Espinosa2011} used unsupervised clustering techniques to investigate expression of 22 genes of this pathway.  
They found that HGSOC cases formed two separate clusters; the cluster with high expression of CASP3, XIAP, NFKB1, FAS and GSK3B was linked with better outcomes.

An integrative network approach has identified two distinct subtypes of ovarian cancer, distinguished by differences in expression of genes related to transcription factors which influence angiogenesis \cite{Glass2015}.  The authors suggest that the regulatory networks highlighted by their analysis can lead to targetting with specific drugs.

A recent pre-publication develops a pipeline for sub-type prediction with the potential to be used clinically \cite{Talhouk2020}.
The NanoString gene expression platform is used.  Their strategy was to focus on small set of 513 genes (vs $\sim$21,000 protein coding genes) known from the literature to have relevance to subtyping.
Two approaches were followed by different teams.  ``All Array": used expression array data from 1650 patients across 14 studies evaluating 9 supervised learning algorithms by bootstrap, selecting an AdaBoost -like method. The ``TCGA" team used 434 patients from The Cancer Genome Atlas (TCGA\footnote{I'm always amused by the genetics pun in the naming of this resource!}) evaluating 5 algorithms by cross-validation, selecting a random forest.  ``All Array" had the advantage of more data but needed careful attention to batch effects. A final classifier took a consensus of the two methods based on a minimal gene set (around 40), validated by a leave-one-out (patient level) approach.  Survival analysis was carried out stratified by predicted subtype.

\subsubsection{Treatment response and survival}

As we have seen in section \ref{sec-epithelial-ovarian-cancer}, EOC subtypes are closely linked with survival, never-the-less much research is directed at finding direct links with expression.

The prognostic value of expression signatures is demonstrated by the ``Classification of Ovarian Cancer" (CLOVAR) system \cite{Verhaak2013}.  
Single-sample gene set enrichment analysis (ssGSEA) is used to obtain scores for previously reported expression signatures: differentiated, immunoreative, mesenchymal and proliferative.  
An outcome prediction model based on these signatures gave good survival stratification,, with likelihood ratio 10.8 between above/below median groups on a validation dataset which improved to 15.7 when augmented with BRCA1/2 mutation status and other post-operative clinical factors.

An explicit supervised learning approach can be taken, as in Berchuck \etal, where a hybrid decision tree plus linear discriminant model demonstrated survival stratification in both cross-validation and external validation experiments, based on microarray expression data.   
They found expression of CSFT3, ABCD3, MAL and APMCF1 to be the most influential; in fact the decision tree was almost entirely driven by the first two of these.

Mairinger \etal\ screened 770 immune related genes to identify 11 differentially expressed genes associated with response to platinum treatment.  
They find that expression of HS11B1, DNBT1, CKLF, NUP107, CCL18, LY96, ATG7, SLAMF7, CXCL9 is associated with better survival, while IKBKG and SDHA associate with with poor survival.

The platinum based drug Cisplatin is a key treatment in ovarian cancer, yet tumours often develop resistance, possibly related to the host's immune response.  
Understanding and predicting such resistance is therefore of clinical importance and is addressed by Mairinger \etal\ \cite{Mairinger2019} through expression analysis using the NanoString platform with a panel of 770 immune related genes.   
They find the following genes to be significantly related to platinum resistance: KLRC1, TCF7, CD274, HSD11B1, COLEC12, PDGFC, FCF1, BMI1, TNFRSF9, ATG10, EWSR1.

Finding a clear, robust correlation between gene expression patterns and survival is not straight forward.  
One substantial meta-analysis applied 16 previously published gene sets to two studies from the Gene Expression Omnibus (GEO) (199 samples) which included survival information, and found no significant predictive power \cite{Fekete2012}.  
Although, within the same study, the authors did identify predictive genes, in particular SNGC, MAPT, ESR2 and PGR.

Expression of a single gene -- CD38, which codes for a transmembrane glycoprotein -- has recently been shown to have survival prognostic value on its own \cite{Zhu2020}.
CD38 has a role in the breakdown of nicotinamide adenine dinucleotide ($\mbox{NAD}^+$), and is also a cell marker of lymphocytes.
CD38 is over expressed in myeloma cells, thence its importance as a prognostic marker and a potential drug treatment target.  
It is worth mentioning that the analysis in this paper made use of the Gene Expression Profiling Interactive Analysis (GEPIA) database, which conveniently supports survival analysis with respect of any given gene, against a range of TCGA hosted gene expression databases.

\subsubsection{Association with other risk factors}

Obesity has been shown to negatively impact prognosis in ovarian cancer, possibly through difficulties in matching chemotherapy dose with patient BMI  \cite{Au-Yeung2014}.  A more direct molecular link is proposed by Cuello \etal\ \cite{Cuello2018}.  They used NMF based clustering to demonstrate that expression of genes linked with obesity and lipid metabolism impart poorer progression free survival, independent of known HGSOC driver gene expression.  The link is complex however, there being many confounding non-molecular based reasons for a link between high BMI and HGSOC survival, including later diagnosis and compromised operability.  Some studies (e.g. \cite{Hew2013}) conclude no overall link at all.

%Stratification of HGSOC by immunohistochemistry is addresses Kamble \etal. \cite{Kamble2019}.  They identify a panel of six biomarkers -- TCF21, E-cadherin, PARP1, Slug and AnnexinA2.   These markers allow stratification into categories including Mesenchymal-to-epithelial transition, Homologous recombination repair and Epithelial-to-mesenchymal transition.  This, they argue, allows selection of class-specific inhibitor drugs such as Olaparib and Rucaparib.

\subsection{Matrix factorization for dimensionality reduction}
\label{sec-matrix-factorization-intro}

@@

The three matrix factorization methods considered in this work are described below.  All are forms of dimensionality reduction, but have differing criteria they aim to optimise.  The number of target dimensions (or components, factors, metagenes -- these terms are used somewhat synonymously) is referred to as the rank, $k$.
\subsubsection{Non-negative matrix factorization (NMF)}
\label{sec-nmf}
As the name implies, this is only applicable to matrices with +ve or zero elements, and finds components which are themselves +ve or zero.   This property make the resulting factorization more interpretable since the components are strictly additive.   
NMF has been applied widely in -omics research, originally to gene expression microarrays  and latterly to RNA-seq datasets.  
A particular value of NMF is that it allows for input samples to be directly assigned to one of $k$ clusters.   
The optimization function on which NMF is based can be tuned to achieve a \emph{sparse} factorization result, that is one in which only a small number of elements in the achieved factorization are non-zero, a property which further simplifies interpretation \cite{Brunet2004}.
%
\subsubsection{Independent component analysis (ICA)} 
The goal of ICA is to represent the given matrix with components which are statistically independent of each other.  It was originally proposed to solve the blind source separation problems \cite{Comon1994}. If the input matrix of samples were drawn from a multivariate Gaussian distribution, then the result would be no different to that of PCA.    Where the data is non-Gaussian however, ICA results in components which separate out independent sources of variation.

\subsubsection{Principal component analysis (PCA)}
PCA is a well known technique for dimensionality reduction, based on eigenvalue decomposition.  It is arguably the ideal, efficient solution for data which is multivariate Gaussian distributed -- or can be assumed to be so.  PCA results in components which are ordered by the proportion of total variance they explain, and are mutually orthogonal.

@@

\subsubsection{Determining the optimum number of factors}

@@

\subsubsection{Comparisons dimensionality reduction methods for gene expression analysis}
\label{sec-comparative-dimensionality-studies}

@@

\subsection{Research gaps and questions}
Considering the foregoing review of the literature, there has clearly been much work by large teams of researchers directed at understanding the gene expression patterns in HGSOC, as one would expect given its clinical importance.  
Dimensionality reduction is a huge and active field -- the bread-and-butter of machine learning research. 
As we have seen, conclusions both on the underlying biology and the optimal methodology are far from consistent, however.   Some approaches -- such as the BioBombe reviewed above -- are complex with results likely to be difficult to reproduce.

@@

\section{Methodology}

\subsection{Outline}

@@

\subsubsection{Datasets}
Two gene expression datasets were made available for this work (N refers to number of patients, m refers to number of protein coding genes).
\begin{enumerate}
\item The Cancer Genome Atlas (TCGA) derived, N=374 patiens, m=19,601 genes, with metadata on survival
\item Australian Ovarian Cancer Study (AOCS) \cite{Patch2015}, N=80, m=19,730, with metadata on survival, cellularity and additional genomic features. Several of the studies reviewed earlier also use data from the Australian Ovarian Cancer Study (AOCS): \cite{Patch2015,Ewing2020,Cuello2018,Au-Yeung2014}.
\end{enumerate}

@@

\subsubsection{Methods}
An overview of the methodology adopted in this work is shown in figure \ref{fig-method-overview}, consisting of:

@@

\subsubsection{Tools}
\begin{itemize}
\itemsep -0.5em 
\item DELL XPS 13, 16GB RAM, running Linux Mint version v19.
\item Python v3.6.8
\item PyCharm Integrated Development Environment, v2020.1.3
\item Numpy for high performance matrix manipulation, v1.18.15
\item Pandas for data frame handing, v1.0.5
\item Matplotlib and Seaborn for general plotting, v3.2.2 and v0.10.1
\item SciPy for least-square solutions and statistical tests, v1.5.0
\item Scikit-learn for matrix factorization and k-means clustering, v0.23.1
\item GOATOOLS package for GEA against GO, v1.0.6
\item Lifelines package for survival analysis, v0.24.16.
\item Github for source control.
\end{itemize}

\subsection{Syncrhonisation of gene sets accross AOCS and TCGA}
\label{sec:gene-set-intersection}

In order to allow factorizations found in the TCGA dataset to be applied to the AOCS dataset it is necessary (or at least convenient) to synchronize the set of genes over which the expression matrices are defined.   As provided to this project, the TCGA and AOCS datasets cover 19,610 and 19,730 protein coding genes respectively, 19,566 of which are common to both (according to ENSG encodings).  Thus, both datasets were pruned to the 19,566 intersection set and ordered consistently.

\subsection{Matrix factorization computation}
One of the aims of this work is to compare the efficacy of three methods of dimensionality reduction -- NMF, ICA and PCA -- as explained in the introduction.     
Some of the key algorithm hyper parameters of each method were investigated and tuned with respect to reconstruction accuracy, specifically the root-mean square (RMS) difference between $X$ and $W H$. The following parameters were explored in each case:
\begin{description}
\item[NMF:] Parameters {\tt max\_iter} (algorithm iterations) and {\tt tol} (tolerance of convergence) were optimised for good accuracy and acceptable execution time.   
{\tt max\_iter} = 5000 and {\tt tol} = 0.01 was used.
Other parameters of interest are {\tt alpha} (multiplier for regulation term) and {\tt l1\_ratio} (multiplier for L1 regularization, when {\tt alpha} $> 0$).  L1 regularization  favours elements being precisely zero, whereas L2 regularization will encourage them to be small.  These parameters have not been explored in this project, and the default {\tt alpha = 0} (no regularization) was used.
\item[ICA:]  Parameters {\tt max\_iter} and {\tt tol} were investigated as for NMF above. 
{\tt max\_iter} = 5000 and {\tt tol} = $10^{-5}$ was used.
Additionally, options for the entropy function which forms the basis of the optimiziation were investigated.
\item[PCA:] This is in principle a deterministic algorithm based on eigenvector decomposition.  However, {\tt sklearn.decomposition.PCA} uses a more efficient 'randomized' algorithm when the given matrix is larger than 500 in both dimensions.  Thus in our use case PCA is seen to have (slightly) stochastic behaviour.
\end{description}

\subsection{Metagene extraction by cluster coherence under bootstrap resampling}
\label{sec-metagene-selection}
Deciding on the number of components (factors, metagenes) to extract -- that is the factorization rank -- is key.  Taking more components results in more accurate representation of the observed expression matrix and provides more avenues to explore the underlying biology.  However, it is important that the metagenes are \emph{stable}, that is that they have reliable meaning when transfered to other datasets.   There are two sources of variation or instability to consider.

@@

\subsection{Determining metagene similarity by Jaccard index}

It is pertinent to ask ``how similar are the 11 metagenes which emerge from the foregoing factorization and cluster analysis?''.  We expect those coming from a single factorization method to be distinct by construction.  But perhaps ICA and PCA identified similar components.  Similarity with NMF derived components is possible but less likely due to the positivity constraint.  A possible approach to measuring similarity might be to treat the components as vectors and determine the angle between them, via the scalar or ``dot'' product.  But in such a high dimension space (19,566) any pair of vectors will be very close to $90^\circ$.

An alternative is to consider the intersection of the gene sets which each metagene highlights.  In the GEA which follows, candidate genes will be identified for each meatagene.  A standard measure of set similarity is the Jaccard index, (or similarity), defined on a pair of sets as:

\[
   \mbox{Jaccard}(A, B) = \frac{A \cap B}{A \cup B}.
\]
The measure lies between 0 (no similarity) and 1 (identical), and is symmetric.  Jaccard similarity was thus calculated for all pairs of the 11 genes sets and visualised as a square heatmap.

\subsection{Gene enrichment analysis (GEA)}
\label{sec-gea}
The metagenes extracted by the above described factorization and clustering process provide valuable information into which genes vary in expression in the study samples; thus in our case, which genes are influential in the dominant expression signal in HGSOC.  
In order to gain insights into what biological processes are involved, GEA against the Gene Ontology (GO) was carried out for each of the 11 metagenes individually.

@@

\subsection{Transfer of learned metagenes to a novel dataset}
\label{sec-transfer-to-novel}

@@

\subsection{Reconciling computed metasamples and metadata}
\label{sec-reconciling-metasamples-metadata}

To carry out patient level analysis -- survival analysis, heatmaps and genome feature correlation -- as described in the following sections, it is necessary to compile tables (Pandas DataFrames) which bring together per-patient metasamples relating to each of the selected metagenes, with associated metadata -- such as cellularity and survival information.   The metagenes always derive from the TCGA dataset (because of its larger size as explained in section \ref{sec-metagene-selection}), but we wish to apply these metagenes as metasamples to the expression data and associated metadata of either the TCGA or AOCS datasets.   
The mathematics of transferring metagenes across datasets has been described above.  Care is required to ensure that expression matrices are aligned correctly with metagenes with respect to their genes, and that metadata is aligned correctly with metasamples with respect to patient identifiers.
Note that the same transferring approach is taken when metasamples are required for TCGA, even though we could in that case obtain the $H$ matrix directly from the factorization.

\subsection{Survival analysis}
\label{sec-survival-analysis}

Survival analysis was performed to investigate whether the metasamples derived from the selected metagenes correlate with patient survival.  Overall  survival (OS) data is available for both the TCGA and AOCS datsets.  Additionally, progression free survival (PFS) data is available for AOCS.   As described above,  metagenes were obtained by factorization on the TCGA dataset only.  Application to TCGA (for OS) thus represents an in-sample test, while application to AOCS (for OS and PFS) is a more exacting out-of-sample test. There are thus three analyses to consider: 
1) TCGA$\rightarrow$TCGA (OS) meaning TCGA derived metagenes are tested for OS on the TCGA dataset, 2) TCGA$\rightarrow$AOCS (OS) meaning TCGA derived metagenes are tested for OS on the AOCS dataset, and 3) TCGA$\rightarrow$AOCS (PFS)... as 2) but for PFS.

Analysis was performed using the Python {\tt Lifelines} package \cite{Davidson-Pilon2020}.  
%Metasamples ($H$ matrices) were calculated from the metagenes and expression matrix as explained in section \ref{sec-transfer-to-novel} below.  
Metasample values were binarized to 0, 1 by thresholding at the median value.   Kaplan-Meier plots were made in respect of derived metasample for each of the three analyses -- making $11 \times 3$ plots each with two survival curves with 95\% confidence intervals.  A hazard ratio (HR)
was calculated for each analysis by fitting Cox's proportional hazards model with p-value relating to the hypothesis that HR is significantly different to 1.0.

\subsection{Investigating correlation between metasamples and genomic features}
\label{sec-genomic-features}

@@

\subsection{Investigating batch effects by analysis of a combined dataset}

@@

\subsection{Codebase and plotting conventions}

The described methodology was implemented in Python.
%except for the KEGG GEA which was implemented in R.
All code is available in a github repository: \url{https://github.com/ipoole/HgsocTromics}.  Good software engineering practice has been followed, with an object-oriented design, frequent commits and unit testing.  A base class is used to present a uniform interface for the three factorization methods.  Intermediate results are cached to file with lazy computation for efficient working without imposing evaluation order dependency.  Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 65 tests executes in around 45 seconds.   The fully automated (except for rank selection) analysis pipeline takes approximately three hours from scratch, the majority of that time being taken for the repeated factorizations over the range of considered ranks.  The total codebase is approximately 2,900 lines of Python.

All plots are generated in vector graphics pdf format to ensure smooth scaling to any resolution.  A consistent colour scheme of \textcolor{blue}{blue}, \textcolor{orange}{orange} and \textcolor{olive}{green} is used for plots relating to NMF, ICA and PCA respectively.  In figures metagenes / metasamples are consistently referred to by, for example, ``NMF-2-of-3" -- meaning the 2nd component of the rank $k=3$ NMF factorization.  In text this is shortened to ``NMF-2'' where the rank is understood.

@@

\section{Results}

% \subsection{Investigation of optimal parameters}
\subsection{Consideration of sampling error is crucial to finding robust metagene signals}

@@

\subsection{Jaccard similarity shows metagenes identify intersecting gene sets in some cases}

The Jaccard heatmap of figure \ref{fig-component-jaccard-similarities} shows that some pairs of metagenes identify overlapping sets of genes, in particular the pairs (ICA-1, PCA-2: J=0.61) and (ICA-5, PCA-3: J=0.49).  As expected however, there is very little similarity between components \emph{within} a factorization method.

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.4]{\plotspath{TCGA_OV_VST/GeneEnrichment/component_jaccard_heatmap.pdf}}
\end{center}
\caption{Heatmap of Jaccard similarities between the candidate genes identified by the 11 components.}
\label{fig-component-jaccard-similarities}
\end{figure}

\subsection{Metagenes highlight genes which are enriched for biological processes}

@@

\subsection{Association of unsupervised gene expression patterns with patient survival is inconclusive}

Kaplan-Meier overall survival (OS) plots stratified by each metasample component are shown for the for the AOCS dataset in figure \ref{fig-kaplan-meier-os-TCGA-AOCS} and for the TCGA dataset in figure \ref{fig-kaplan-meier-os-TCGA-TCGA} (appendix).  Plots for progression free survival (PFS) on AOCS are also in the appendix, figure \ref{fig-kaplan-meier-pfs-TCGA-AOCS}.   All plots show 95\% confidence intervals and hazard ratio (HR) with associated p-value.

All of these results are summarised in figure \ref{fig-summary-survival-plots}, which brings together the three sets of results -- TCGA (OS), AOCS (OS) and AOCS (PFS) --  with respect to the 11 metasample components.  $\log_2{\mbox{HR}}$ is used in order that the \emph{sense} of the survival impact can be readily appreciated.    If a metasample component has a robust correlation with survival, then we expect the p-value for all three sets of results to show significance \emph{and} for the effects to have the same sense -- be in the same direction.  None of the 11 components pass this test.

The component with the largest observed effect on survival is PCA-1, showing a reasonably consistent hazard ratio of around 1.3 ($2^{0.4}$) across the three experiments.  This has $p < 0.05$ in the larger TCGA dataset, but not in the smaller AOCS dataset.

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.4]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_AOCS.pdf}} 
\caption{Kaplan-Meier plots for TCGA $\rightarrow$ AOCS for overall survival (OS).}
\label{fig-kaplan-meier-os-TCGA-AOCS}
\end{center}
\end{figure}

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.4]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_NMF.pdf}} \\
\includegraphics[scale=0.4]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_ICA.pdf}} \\
\includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_PCA.pdf}}
\caption{Visual summary of survival analysis as applied to TCGA(OS), AOCS(OS) and AOCS(PFS).  Plots are divided by factorization method.  Bar heights show $\log_2(\mbox{HR})$ with p-value also shown.}
\label{fig-summary-survival-plots}
\end{center}
\end{figure}

\subsection{Metasamples correlate with some genomic features}

@@

\subsection{Integrative analysis of results}
\label{sec-integrative-analysis}

@@

\subsection{Matrix factorization discovers and effectively removes batch effects}

@@

\section{Discussion}

@@

\subsubsection*{What underlying biological processes are at play in HGSOC as seen through the patterns of gene expression? }

Biological processes identified by GEA have been set out in section \ref{sec-gea}.   Processes relating to the ECM and extracellular organisation may reflect cell type heterogeneity -- perhaps within the tumour its self, or due to sampling of adjacent normal tissue.   Processes of negative regulation of angiogenesis could reflect tumour growth and therefore be of prognostic value.

These observations are based on GEA on the GO only.  Possibly more insights could have been gained by GEA against KEGG\footnote{\href{https://www.genome.jp/kegg/pathway.html}{Kyoto Encyclopedia of Genes and Genomes}} Pathways. (This was begun but had to be dropped due to shortage of time).

\subsubsection*{How do the patterns of gene expression uncovered relate to genome level features, such as those marking genome instability?}

The integrative analysis set out in section \ref{sec-integrative-analysis} suggests that  Cellularity is correlated to the processes of the the ECM, reflecting heterogeneity of tissue type.   Of course, Cellularity is not really a biological characteristic -- it is a technical artefact of the tumour biopsy process, and so we might conclude that signatures relating to ECM might also be considered as technical artefacts.  We found that whole genome doubling (WGD) and Cellularity correlate together with three components even though they are not themselves directly correlated -- something which warrants further investigation.

Weak evidence of a link between homologous repair deficiency (HRD) and RNA-binding was found.   Being the site of protein synthesis, ribosomes will be associated with most of what goes on in a the cell.   One paper suggests a link : ``a comprehensive review of the literature reveals a role for conventional DNA repair proteins in ribosome biogenesis, and conversely, ribosome biogenesis proteins in DNA repair" \cite{Ogawa2018}.   Seek and ye shall find!

A note of caution: these results are based on testing $6 \times 11 = 66$ hypotheses at a significance level of $< 0.01$, so there will likely be at least one false discovery.  A significance level of 0.001 would have been more appropriate.

Expression signature correlations with mutational load and  CNV load have not been detected.

\subsubsection*{Do MF methods yield factors which are predictive of patient survival?}

We found no convincing evidence of this, for an individual factor (metasample).  Some components were predictive in one or other of the two datasets, but not both, which is unconvincing.  The factor for which GEA indicated regulation of angiogenesis (NMF-1), and thence might have prognostic value, performed poorly in survival analysis.  So overall a negative result.

Is there a signal for survival to be found in these data?  We have noted that according to \cite{Zhu2020}, the single gene CD38 is highly predictive of survival in EOC.  Does this hold for the TCGA and AOCS datasets we are working with?  This is easy to check, running survival analysis stratified by a hand-crafted metagene, setting the CD38 element to 1, all others to 0.   This was applied as in section \ref{sec-survival-analysis} to TCGA (OS),  AOCS (OS) and AOCS (PFS) -- see figure \ref{fig-cd38 survival}.   The effects are in a consistent direction across the three tests, with high CD38 expression predicting favourable survival.  This is highly significant in the TCGA dataset and $p < 0.05 $ in the smaller AOCS.  So there is a signal, but it did not emerge unbidden from matrix factorization.  (CD38 is in fact highlighted by component ICA-5, see figure \ref{fig-ICA-5-5-go-lineage}, but must be diluted by other genes).

>>>>

\documentclass[tikz, 11pt,a4paper,oneside,fleqn, draft]{article}

\usepackage[margin=2.5cm]{geometry}

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{setspace}
\usepackage{enumitem}
\setlist[description]{leftmargin=\parindent,labelindent=\parindent}

@@

\title{Accounting for sampling error in when applying matrix factorization to gene expression data}

\author{Ian Poole and Ailith Ewing}
\date{January 2021}

@@

\subsection{- - Epithelial Ovarian Cancer}
\label{sec-epithelial-ovarian-cancer}

@@

\subsection{- - Patterns of gene expression in HGSOC}
\label{sec-review-of-gene-expression}
Identifying the patterns -- or ``signatures'' -- of gene expression has been an active research area, for the purposes of understanding the carcinogenesis pathway, subtype classification, predicting survival prognosis, predicting treatment response and understanding the causative mechanism of risk factors; there is much cross-over between these areas, however.

\subsection{+ Matrix factorization for dimensionality reduction}
\label{sec-matrix-factorization-intro}

@@

The three matrix factorization methods considered in this work are described below.  All are forms of dimensionality reduction, but have differing criteria they aim to optimise.  The number of target dimensions (or components, factors, metagenes -- these terms are used somewhat synonymously) is referred to as the rank, $k$.
\subsubsection{+ Non-negative matrix factorization (NMF)}
\label{sec-nmf}
As the name implies, this is only applicable to matrices with +ve or zero elements, and finds components which are themselves +ve or zero.   This property make the resulting factorization more interpretable since the components are strictly additive.   
NMF has been applied widely in -omics research, originally to gene expression microarrays  and latterly to RNA-seq datasets.  
A particular value of NMF is that it allows for input samples to be directly assigned to one of $k$ clusters.   
The optimization function on which NMF is based can be tuned to achieve a \emph{sparse} factorization result, that is one in which only a small number of elements in the achieved factorization are non-zero, a property which further simplifies interpretation \cite{Brunet2004}.
%
\subsubsection{+ Independent component analysis (ICA)} 
The goal of ICA is to represent the given matrix with components which are statistically independent of each other.  It was originally proposed to solve the blind source separation problems \cite{Comon1994}. If the input matrix of samples were drawn from a multivariate Gaussian distribution, then the result would be no different to that of PCA.    Where the data is non-Gaussian however, ICA results in components which separate out independent sources of variation.

\subsubsection{+ Principal component analysis (PCA)}
PCA is a well known technique for dimensionality reduction, based on eigenvalue decomposition.  It is arguably the ideal, efficient solution for data which is multivariate Gaussian distributed -- or can be assumed to be so.  PCA results in components which are ordered by the proportion of total variance they explain, and are mutually orthogonal.

@@

\subsubsection{- - Determining the optimum number of factors}

@@

\subsubsection{- - Comparisons dimensionality reduction methods for gene expression analysis}
\label{sec-comparative-dimensionality-studies}

@@

\subsection{- - Research gaps and questions}
Considering the foregoing review of the literature, there has clearly been much work by large teams of researchers directed at understanding the gene expression patterns in HGSOC, as one would expect given its clinical importance.  
Dimensionality reduction is a huge and active field -- the bread-and-butter of machine learning research. 
As we have seen, conclusions both on the underlying biology and the optimal methodology are far from consistent, however.   Some approaches -- such as the BioBombe reviewed above -- are complex with results likely to be difficult to reproduce.

@@

\section{- - Methodology}

\subsection{- - Outline}

@@

\subsubsection{+ Datasets}
Two gene expression datasets were made available for this work (N refers to number of patients, m refers to number of protein coding genes).
\begin{enumerate}
\item The Cancer Genome Atlas (TCGA) derived, N=374 patiens, m=19,601 genes, with metadata on survival
\item Australian Ovarian Cancer Study (AOCS) \cite{Patch2015}, N=80, m=19,730, with metadata on survival, cellularity and additional genomic features. Several of the studies reviewed earlier also use data from the Australian Ovarian Cancer Study (AOCS): \cite{Patch2015,Ewing2020,Cuello2018,Au-Yeung2014}.
\end{enumerate}

@@

\subsubsection{- - Methods}
An overview of the methodology adopted in this work is shown in figure \ref{fig-method-overview}, consisting of:

@@

\subsubsection{- - Tools}
\begin{itemize}
\itemsep -0.5em 
\item DELL XPS 13, 16GB RAM, running Linux Mint version v19.
\item Python v3.6.8
\item PyCharm Integrated Development Environment, v2020.1.3
\item Numpy for high performance matrix manipulation, v1.18.15
\item Pandas for data frame handing, v1.0.5
\item Matplotlib and Seaborn for general plotting, v3.2.2 and v0.10.1
\item SciPy for least-square solutions and statistical tests, v1.5.0
\item Scikit-learn for matrix factorization and k-means clustering, v0.23.1
\item GOATOOLS package for GEA against GO, v1.0.6
\item Lifelines package for survival analysis, v0.24.16.
\item Github for source control.
\end{itemize}

\subsection{- - Matrix factorization computation}
One of the aims of this work is to compare the efficacy of three methods of dimensionality reduction -- NMF, ICA and PCA -- as explained in the introduction.     
Some of the key algorithm hyper parameters of each method were investigated and tuned with respect to reconstruction accuracy, specifically the root-mean square (RMS) difference between $X$ and $W H$. The following parameters were explored in each case:
\begin{description}
\item[NMF:] Parameters {\tt max\_iter} (algorithm iterations) and {\tt tol} (tolerance of convergence) were optimised for good accuracy and acceptable execution time.   
{\tt max\_iter} = 5000 and {\tt tol} = 0.01 was used.
Other parameters of interest are {\tt alpha} (multiplier for regulation term) and {\tt l1\_ratio} (multiplier for L1 regularization, when {\tt alpha} $> 0$).  L1 regularization  favours elements being precisely zero, whereas L2 regularization will encourage them to be small.  These parameters have not been explored in this project, and the default {\tt alpha = 0} (no regularization) was used.
\item[ICA:]  Parameters {\tt max\_iter} and {\tt tol} were investigated as for NMF above. 
{\tt max\_iter} = 5000 and {\tt tol} = $10^{-5}$ was used.
Additionally, options for the entropy function which forms the basis of the optimiziation were investigated.
\item[PCA:] This is in principle a deterministic algorithm based on eigenvector decomposition.  However, {\tt sklearn.decomposition.PCA} uses a more efficient 'randomized' algorithm when the given matrix is larger than 500 in both dimensions.  Thus in our use case PCA is seen to have (slightly) stochastic behaviour.
\end{description}

\subsection{+ Metagene extraction by cluster coherence under bootstrap resampling}
\label{sec-metagene-selection}
Deciding on the number of components (factors, metagenes) to extract -- that is the factorization rank -- is key.  Taking more components results in more accurate representation of the observed expression matrix and provides more avenues to explore the underlying biology.  However, it is important that the metagenes are \emph{stable}, that is that they have reliable meaning when transfered to other datasets.   There are two sources of variation or instability to consider.

@@

\subsection{- - Gene enrichment analysis (GEA)}
\label{sec-gea}
The metagenes extracted by the above described factorization and clustering process provide valuable information into which genes vary in expression in the study samples; thus in our case, which genes are influential in the dominant expression signal in HGSOC.  
In order to gain insights into what biological processes are involved, GEA against the Gene Ontology (GO) was carried out for each of the 11 metagenes individually.

@@

\subsection{- - Transfer of learned metagenes to a novel dataset}
\label{sec-transfer-to-novel}

@@

\subsection{- - Investigating correlation between metasamples and genomic features}
\label{sec-genomic-features}

@@

\subsection{- - Investigating batch effects by analysis of a combined dataset}

@@

@@

\section{- - Results}

% \subsection{Investigation of optimal parameters}
\subsection{+ Consideration of sampling error is crucial to finding robust metagene signals}

@@

\subsection{- - Metagenes highlight genes which are enriched for biological processes}

@@

\subsection{- - Metasamples correlate with some genomic features}

@@

\subsection{+ Integrative analysis of results}
\label{sec-integrative-analysis}

@@

\subsection{- - Matrix factorization discovers and effectively removes batch effects}

@@

\section{- - Discussion}

@@

\subsubsection*{- - How should MF methods be applied to achieve robust results on datasets of limited size?}
